### 2015-06-18 chapters/ipa_meets_unicode.tex
update treatment of middle tilde

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 13 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}.}

\begin{itemize}
	\item Basic Latin (27 characters): \newline < a~b~c~d~e~f~h~i~j~k~l~m~n~o~p~q~r~s~t~u~v~w~x~y~z~.~| >
	\item Latin-1 Supplement (4 characters): \newline < æ ç ð ø >
	\item Latin Extended-A (3 characters): \newline < ħ ŋ œ >
	\item Latin Extended-B (4 characters): \newline < ǀ ǁ ǂ ǃ >
	\item IPA Extensions (68 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɫ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩ >
	\item Combining Diacritical Marks (29 characters): \newline < {\large \  ̃\ \ ̆\ \ ̈\ ̚\ \ ̘\ \ ̙\ \ \ ̜\ \ ̝\ \ ̞\ \ ̟\ \ ̠\ \ ̤\ \ ̥\ \ ̩\ \ ̪\ \ ̬\ \ ̯\ \ ̰\ \ ̹\ \ ̺\ \ ̻\ \ ̼\ \ ̽\ \ ͜\  ̋\ \ ́\ \ ̄\ \ ̀\ \ ̏\ } >
	\item Greek and Coptic (3 characters): \newline < β θ χ >
%	\item Phonetic Extensions (2 characters): \charis{ᴅ ᴴ} 
	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
    \item General Punctuation (2 characters): \newline < ‖ \charis{‿} >
	\item Superscripts and Subscripts (1 character): \newline < ⁿ > 
	\item Arrows (4 characters): \newline < ↑ ↓ ↗ ↘ >
	\item Latin Extended-C (1 character): \newline < \charis{ⱱ} >
\end{itemize}

@@

   \item The dental click <ǀ> and the indication of a minor group break <|>
           (proposed to be used for separating feet) look almost the same in
           most fonts. For a proper encoding, the \textsc{latin letter dental
           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
           should be strictly distinguished.
   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
           which according to the IPA is a major group break (by intonation), to be
           encoded by \textsc{double vertical line} at \uni{2016}.
   \item The marking of primary stress < ˈ > looks like an apostrophe, and
           is often typed with the same symbol as the ejective < ʼ >. For a
           proper encoding, these two symbols should be typed as 
           \textsc{modifier letter vertical line} at \uni{02C8} and
           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively.

@@

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The placement of diacritics
\end{itemize}

@@

Third, for marking marking of voiceless pronunciation of voiced segments the IPA
uses the ring diacritic. Originally, the ring should be placed below the 
base character, like in <m̥>, using the \textsc{combining ring below} at \uni{0325}. 
However, in letters with long descenders the IPA also allows to put the ring 
above the base, like in <ŋ̊>, using the \textsc{combining ring above} at 
\uni{030A}. Yet, proper font design does not have any problem with rendering 
the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
propose to standardize on the ring below. As a principle, for strict IPA 
encoding only one options should be allowed.

>>>>

As explained in Section~\ref{pitfall-blocks}, the Unicode Standard code space is
subdivided into character blocks, which generally encode characters from a
single script. However, as is illustrated by the IPA, characters that form a
single writing system may be dispersed across several different character
blocks. With its diverse collection of symbols from various scripts and
diacritics, the IPA is spread across 12 blocks in the Unicode
Standard:\footnote{This number of blocks depends on whether only IPA-sanctioned
symbols are counted or if the phonetic symbols commonly found in the literature
are also included, see~\cite[Appendix~C]{Moran2012}.}

\begin{itemize}
	\item Basic Latin (27 characters): \newline < a b c d e f h i j k l m n o p q r s t u v w x y z~.~| >
	\item Latin-1 Supplement (4 characters): \newline < æ ç ð ø >
	\item Latin Extended-A (3 characters): \newline < ħ ŋ œ >
	\item Latin Extended-B (4 characters): \newline < ǀ ǁ ǂ ǃ >
	\item IPA Extensions (67 characters): \newline < ɐ ɑ ɒ ɓ ɔ ɕ ɖ ɗ ɘ ə ɛ ɜ ɞ ɟ ɠ ɡ ɢ ɣ ɤ ɥ ɦ ɧ ɨ ɪ ɬ ɭ ɮ ɯ ɰ ɱ ɲ ɳ ɴ ɵ ɶ ɸ ɹ ɺ ɻ ɽ ɾ ʀ ʁ ʂ ʃ ʄ ʈ ʉ ʊ ʋ ʌ ʍ ʎ ʏ ʐ ʑ ʒ ʔ ʕ ʘ ʙ ʛ ʜ ʝ ʟ ʡ ʢ > 
	\item Spacing Modifier Letters (17 characters): \newline < ʰ ʲ ʷ ʼ ˈ ˌ ː ˑ ˞ ˠ ˡ ˤ ˥ ˦ ˧ ˨ ˩ >
	\item Combining Diacritical Marks (29 characters): \newline < {\large \  ̃\ \ ̆\ \ ̈\ ̚\ \ ̘\ \ ̙\ \ \ ̜\ \ ̝\ \ ̞\ \ ̟\ \ ̠\ \ ̤\ \ ̥\ \ ̩\ \ ̪\ \ ̬\ \ ̯\ \ ̰\ \ ̹\ \ ̺\ \ ̻\ \ ̼\ \ ̽\ \ ͜\  ̋\ \ ́\ \ ̄\ \ ̀\ \ ̏\ } >
	\item Greek and Coptic (3 characters): \newline < β θ χ >
%	\item Phonetic Extensions (2 characters): \charis{ᴅ ᴴ} 
%	\item Phonetic Extensions Supplement (1 characters): \newline < \charis{ᶑ} >
    \item General Punctuation (2 characters): \newline < ‖ \charis{‿} >
	\item Superscripts and Subscripts (1 character): \newline < ⁿ > 
	\item Arrows (4 characters): \newline < ↑ ↓ ↗ ↘ >
	\item Latin Extended-C (1 character): \newline < \charis{ⱱ} >
\end{itemize}

@@

   \item The dental click <ǀ> and the indication of a minor group break <|>
           (proposed to be used for separating feet) look almost the same in
           most fonts. For a proper encoding, the \textsc{latin letter dental
           click} at \uni{01C0} and the \textsc{vertical line }at \uni{007C} 
           should be strictly distinguished.
   \item Similarly, the alveolar lateral click <ǁ>	should be encoded with a
           \textsc{latin letter lateral click} at \uni{01C1}, different from <‖>, 
           which according to the IPA is a major group break (by intonation), to be
           encoded by \textsc{double vertical line} at \uni{2016}.
   \item The marking of primary stress < ˈ > looks like an apostrophe, and
           is often typed with the same symbol as the ejective < ʼ >. For a
           proper encoding, these two symbols should be typed as 
           \textsc{modifier letter vertical line} at \uni{02C8} and
           \textsc{modifier letter apostrophe} at \uni{02BC}, respectively. 
   \item There are two different ``dashed''-l characters in IPA, namely the <ɫ> 	      \textsc{latin small letter l with middle tilde} at \uni{026B} and the <ɬ>	      \textsc{latin small letter l with belt} at \uni{026C}. These of course
          look highly similar, although they are different sounds. As a solution, 
          we will actually propose to not use the middle tilde at all 
          (see Section~\ref{pitfall-multiple-options-ipa}).

@@

\begin{itemize}
  \item The marking of tone
  \item The marking of <g>
  \item The marking of velarization ad pharyngealization
  \item The placement of diacritics
\end{itemize}

@@

Third, the IPA has different markers for velarization <ˠ> and pharyngealization
<ˤ>. Confusingly, there is also a marker for ``velarized or pharyngealized'',
using the <\ \ {\large ̴}> \textsc{combining tilde overlay} at \uni{0334}. The
tilde overlay seems to be extremely rarely used. For strict IPA encoding we
propose to not include the combining tilde overlay at all, but only use the 
other two Spacing Modifiers.

Finally, for marking marking of voiceless pronunciation of voiced segments the IPA
uses the ring diacritic. Originally, the ring should be placed below the 
base character, like in <m̥>, using the \textsc{combining ring below} at \uni{0325}. 
However, in letters with long descenders the IPA also allows to put the ring 
above the base, like in <ŋ̊>, using the \textsc{combining ring above} at 
\uni{030A}. Yet, proper font design does not have any problem with rendering 
the ring below the base character, like in <ŋ̥>, so for strict IPA encoding we 
propose to standardize on the ring below. As a principle, for strict IPA 
encoding only one options should be allowed.
